 

### 2023-07-18 markdown/Thesis.tex
until batch correction done

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
2cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] { \expandafter\origfigure\expandafter[H] } { \endorigfigure }
\usepackage{sectsty}
\usepackage{paralist}
\usepackage{setspace}\spacing{1.5}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{dcolumn}
\usepackage{natbib}\bibliographystyle{agsm}
\usepackage[nottoc, numbib]{tocbibind}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\Large
\doublespacing
{\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
presenting ProteoScanR & Proteomics Workbench

@@

Main Supervisor: 
Dr. DI(FH) Gerhard Duernberger

External Supervison:
Assoc. Prof. Ujjwal Neogi, M.Sc. PhD

Second examiner:
FH-Prof. Dr. Alexandra Graf

@@

For testing and development of the ProteoScanR pipeline the following
materials and methods were performed. For the sample preparation
protocol see appendix.

@@

Macrophages from 12 healthy donors and 12 HIV patients with metabolic
syndrome and 12 HIV patients without metabolic syndrome. Cell sorting
was performed with FACSAria Fusion (BD Biosciences).

@@

@@

Samples of test dataset were labeled with 12-plex TMTpro (Thermo Fisher)
in anhydrous acetonitril (ACN) solution.

@@

@@

Lenovo Laptop X360 with: Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz 8 GB
RAM

@@

{\centering \includegraphics{Thesis_files/figure-latex/data_processing_pipeline_flowchart_vertical-1}

@@

\caption{Flowchart presenting the processing by ProteoScanR}\label{fig:data_processing_pipeline_flowchart_vertical}
 \endfigure\egroup

@@

In order to streamline the analysis of multiple experiments, object
oriented programming was applied. The approach in R is to create a so
called Q-feature object, which contains all variables and metadata in a
hierarchical structure. The structure enables sub setting for further
analysis \citep{Vanderaa2021}.

@@

The next step for quality control is the exclusion of samples with a
high false discovery rate (FDR). When applying multiple statistical
testing (e.g.~t-Test) the obtained p-values can be considered as biased,
because the probability to observe a significant will iteratively
increase with each test performed. Corrections in statistics are an
approach to compensate for the multiplicity of testing. There are many
ways to do this compensation like the Bonferroni method or
Benjamini-Hochberg`s FDR. In Mass spectrometry the common ``way to go''
is calculating a false discovery rate, by dividing false PSMs (=hit of
the decoy database) through the total number of PSMs above the
peptide-spectrum matching score. The peptides spectrum matching score is
defined as \(-10 * log(p)\). Whereas the p-value is defined that the hit
is done by chance. The calculation of the score is highly dependent on
the data acquisition method used. MaxQuant uses Andromeda, an integrated
search engine. Proteome Discoverer from Thermo Fisher utilizes different
engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot
and Andromeda showed similar performance when comparing FDR values as a
function of coverage. However the observed performance can be lower when
dealing with a decreased coverage \citep{Cox2011}.The threshold for
accepting an FDR of an individual PSM is described as q-value.

@@

In data science, aggregation refers to a row-wise operation that merges
data based on a particular column using a specific function. In the
context of processing from PSMs to unique peptides, the desired column
is the PSM sequence. To account for different distributions across
multiple assays, the median of the channel is used as the function to
aggregate multiple matches into one.

@@

Sample size is often a limiting factor in hypothesis testing. A strict
quality control and the fact that TMT reagents are only available up to
18-plex can reduce the number of observed samples below the critical
threshold, leading to an early end of analyses. To overcome this
limitation, the provided software is capable of processing multiple runs
simultaneously, allowing for testing of multiple batches and increasing
the number of samples that can be included in the analysis. For further
processing the runs are joined in one object at the peptide level.

@@

Columns which do not meet the desired intensity can be filtered by a
threshold set on the RI. The median RI can also be used to check if an
entire channel has a lower detection level. This can be due two reasons.
One is the expression level of the given proteins in a cell. Meaning,
that the expression of the observed cell type is simply lower than the
other type. Another one could be a spillage of TMT detection in other
channels due incorrect or missing correction of the TMT isotopes. The
correction values can be found on the vendor`s webpage (in this case
Thermo Fisher) and correction needs to be set up in MaxQuant.

@@

Depending having a bulk sample or single-cell sample choosing a minimum
of observed peptides and a cutoff value for the CV, changes the level of
confidence in the peptide data. The coefficient of variation of a
peptide is considered as the ratio of the standard deviation to the mean
and describes the relationship of the observed peptide signal over
multiple proteins (=razor proteins). Peptides having a high coefficient
of variation over many razor proteins are considered as noise and need
to be filtered out before statistical analysis.

@@

Data transformation applies a function to each value of a matrix or
array, so that: \[
y_i = f(x_i)
\] Depending on the distribution of the values in the observed
expression set, different transformations can be applied to fulfill
assumptions for statistical testing. In the shiny application, various
procedures are implemented and can be further expanded upon request from
the user. The macrophage analysis done by Specht et al.~mentioned in
their SCoPE2 publication \citep{Specht2021} uses the logarithm to the
base 2 to spread a compacted distribution and remove skewness in the
dataset. This transformation was used as a reference when comparing
methods. However, using the logarithm to the base 10 may be an easy way
to recalculate expression data in the mind and will also be facilitated
by other transformation methods such as the boxcox method, which is also
implemented in the application. When observing a wide distribution of
small and large values of the expression set in a histogram, a square
root transformation can help increase the variability of smaller values
and decrease the variability of larger values. Depending on the
distribution, the user has to decide which transformation to consider
and verify that statistical assumptions are met after application with
visualizations such as histograms, qq-plots, and MA-plots. For
statistically inexperienced users, the boxcox transformation can help
with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992},
this method applies a linear model against 1 on the data to determine
the statistical parameter lambda by the maximum likelihood method. The
log-likelihood method takes the 95\% confidence interval for the
parameter lambda and the final lambda is chosen as the value with the
highest log-likelihood value.

@@

The term normalization is ambiguous in data-science and will be
explained briefly in this chapter. Starting with converting the data to
the Z-distribution (Gaussian result), where 0 represents the mean and 1
represents the standard deviation, in vectorization every feature (or
protein) is represented as a vector that points in a specific direction
in the unit sphere, normalization can have different meanings.
Normalization is considered as part of the scaling procedure and ensures
that every feature contributes equally to the statistical model and
hinders large values from biasing the model in a particular direction.
However, this could also have negative consequences for the modeling
procedure, as it may reduce the impact of important features on the
dataset. Depending on the algorithms later used in the data processing,
the method of choice for normalization can change the results
significantly. The K-nearest neighbor (K-NN) clustering algorithm, for
instance relies on the distances between distinct data points by
minimizing variance of the squared euclidean distances. This method will
be applied in the missing value imputation, which will be the next step
applied to the data. The first normalization method available in the
user interface is column-wise median and row-wise mean normalization.
This method was chosen by Specht et al in the SCoPE2 publication 2021
\citep{Specht2021} and worked as a reference during the development
process to benchmark other methods. When dividing each value of an
expression set by the median of the particular column, the procedure is
considered as a column-wise median normalization. Although the
mean-normalization works the same way but takes the mean as the second
variable. In the analysis pipeline both procedures were applied on the
dataset. However when the logarithm as the transformation method was
chosen as transformation method, ProteoScanR automatically switches from
a division to a subtraction, since:
\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\). After applying
this normalization method, each value can be interpreted as a distance
or fraction of the mean or median of the corresponding column or row,
enabling fair comparisons between features and samples. The second
method for normalization included is the CONSTANd method, which was
proven suitable for relative quantification in the field of proteomics
\citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq
data. CONSTANd uses a technique called matrix raking, which employs the
RAS algorithm. The expression set reflects the non-negative real
\((m, n)\) matrix \(A\) were the bi proportional constrained matrix
problem will be solved by finding the \((m, n)\) matrix \(B\) which
equals to \(diag(x) * A *diag(y)\) . Whereas \(x \in \mathbb{R}^{m}\)
and \(y \in \mathbb{R}^{n}\). The solution to this problem involves
finding the row sum of \(B=u_i\) and the column sum of \(B=v_j\), where
\(i\) and \(j\) denote the row and column indices of the matrix,
respectively \citep{Bacharach1965}. In other words, the matrix will be
alternatively manipulated on rows and columns until the mean of both
equals 1. In order to apply the method and yield true results, one has
to consider 3 major assumptions between sample types. This assumption
can be observed in the MA-plot which indicates the differences of two
samples. Given the intensities of two samples, \(R\) and \(G\),
visualized on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\) and
the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all
sample types of identically processed sets is needed before calculation
and the assumptions must be fulfilled.

@@

When analyzing multiple batches simultaneously, it is possible to
encounter situations where a particular protein is not detected in one
of the batches. From a biological perspective, it is highly unlikely
that a cell completely lacks a single protein, as it would imply a
complete loss of function for that protein. Therefore, an intensity
value of 0 is close to impossible for most proteins. In order to handle
this dark space in the expression set, several methods can be chosen in
the application. A variety of function employs replacing the missing
values with the mean or median of the rest of the matrix. This procedure
introduces no bias in the data if the missingness is low, although
differential expression for the sample carrying the missing feature can
not be expected as well. Even a more conservative approach is dropping
the rows for the feature with the missing value in one of the observed
samples. A common practice in proteomics is utilizing the K-nearest
neighbor algorithm (=KNN), which imputes the missing values by a
euclidean metrics of the neighboring values in the columns where the
feature is not missing. \(K\) denotes for the count of neighboring
values of a gene and can be selected upon preference. The euclidean
distance between two points is defined as \(d = \sqrt{a^2 * b^2}\), the
\(K\) nearest neighbors average value will be assigned to the intensity
of the missing feature. Since the average is a statistical metric which
can be biased easily by skewed distributions, normalization is an
important task before performing missing value imputation. KNN was
benchmarked for the regression of protein abundance intensities and
showed an Area under the Curve (AUC) above 0.7 when testing on a human
proteome test set \citep{Lan2013}.

@@

After differential expression the upcoming question could be, in which
context the up or down regulated proteins are. Enriched pathways can be
understood as an over-representation (ORA = over-representation
analysis) of proteins within a specific set. This set can be a pathway
or a selection of common denominators in biological contexts, such as
cell types or oncological factors. The statistical testing to identify
the overabundant proteins of a specific set is similar to a \(\chi^2\)
test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher
test \citep{Sprent2011} for tables with 4 fields in small sample sizes.
Within the list are all truly differentially expressed genes, an
annotation is the identification of a protein in the desired biological
context. The table is expected to be populated by the values as follows:

@@

Under the null hypothesis \(H_{0}\) states that the distribution and
population of the table is random and there is no clear tendency with
significant evidence. Against that the alternative hypothesis \(H_{A}\)
says that there is an underlying tendency the table is populated. This
leads to the test statistics, the observed values (=\(O\)) in the table
are significantly different to the expected
(=\(E=\frac{(A+B \lor C+D)}{N}\) ones. One approach is approximating
\(\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}\) and deriving the
p-value by the area under the density curve of the \(\chi^2\)
distribution \(Q = \sum_{i=1}^{k} Z_{i}^{2}\). In the distribution \(k\)
is defined by the degrees of freedom (=df), what can be obtained by
\(k=(columns-1)*(rows-1)=df\) of the table. Another approach to obtain
an exact p-value would be Fisher`s exact test, which states that the
margins are distributed according to the hypergeometric distributions.
The test statistic leads directly to the p-value with no need for
approximation:
\(p=1-\sum_{i=0}^{A-1}\frac{\binom{B}{i}\binom{D-B}{A+C-i}}{\binom{D}{A+C}}\)
Here D denotes either for the background distribution which can be the
total number of proteins obtained in the experiment, all proteins with
annotation or a custom selected background. The background indicates all
proteins which could be positive tested for in the observed sample. The
above mentioned three options are implemented in the user interface.
After obtaining the p-values for each protein set, the subsequent
critical step involves the correction for multiple testing, considering
the inherent risk of false positive results. Multiple testing correction
methods adjust the significance thresholds to account for the increased
probability of detecting false positives when performing multiple
statistical tests. In this protein set enrichment analysis, the user is
provided with a selection of established correction methods, akin to
those commonly utilized in the field of differential expression
analysis, ensuring robust and reliable interpretation of the results.
Once the enriched protein pathways are determined, they can be visually
represented using bar or dotplots, and a network of the found pathways.
In the network ontologies are depicted as nodes, whereas edges show the
strength of the interaction by thickness corresponding to the number of
associated proteins.

@@

For custom ontologies the protein enrichment analysis is done using the
R bioconductor package piano \citep{Vaeremo2013}. In opposite to the
pathway enrichment, piano maps proteins to a pre-selected gene
collection set, which can be either downloaded from gsea-msigdb.org or
custom made for the particular experiment and explains conditions,
phenotype or other desired properties. For statistical testing the above
explained Fisher exact test is used.

@@

To begin using the application, the first step is to upload the
tab-delimited evidence file and the sample annotation file. Once these
files are uploaded, the application is ready to run with default values.
The default settings are preconfigured for an initial analysis, and any
further adjustments or fine-tuning can be made in subsequent steps of
the analysis process.

@@

\caption{Summary showing the process of the analysis pipeline}\label{fig:ui_summary}
 \endfigure\egroup

After a runtime of 2 minutes on the 4 core laptop the data is ready to
visualize. On the left side of the user interface, users can find the
analysis pipeline settings that can be adjusted. Once the desired
changes are made, they can be applied by clicking the update button.
This allows users to observe the direct impact of the applied changes on
the dataset. In the summary visualization, it may appear cluttered due
to the processing of 36 files. In smaller datasets, the nodes of the
network graph can be more easily distinguished and understood. When
running the application with SCoPE2 \citep{Specht2021} reference
cutoffs, the following results were visualized.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] { \expandafter\origfigure\expandafter[H] } { \endorigfigure }
\usepackage{sectsty}
\usepackage{paralist}
\usepackage{setspace}\spacing{1.5}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{dcolumn}
\usepackage{natbib}\bibliographystyle{agsm}
\usepackage[nottoc, numbib]{tocbibind}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

\Large
\doublespacing
{\bf Design and implementation of analysis pipeline for single cell type proteomics data} 
\\ presenting ProteoScanR \& Proteomics Workbench

@@

\doublespacing
\doublespacing
Main Supervisor: Dr. DI(FH) Gerhard Duernberger \\
External Supervisor: Assoc. Prof. Ujjwal Neogi, M.Sc. PhD \\
Second examiner: FH-Prof. Dr. Alexandra Graf \\

@@

For testing and development of the ProteoScanR pipeline the following materials and methods were applied.

@@

Macrophages from 12 healthy donors and 12 HIV patients with metabolic syndrome and 12 HIV patients without metabolic syndrome were isolated. Cell sorting was performed with FACSAria Fusion (BD Biosciences) by auto fluorescence.

\hypertarget{cells-sorting-protocol}{%
\paragraph{Cells sorting protocol}\label{cells-sorting-protocol}}

@@

\hypertarget{digestion-protocol}{%
\paragraph{Digestion protocol}\label{digestion-protocol}}

@@

Samples of the test data set were labeled with 12-plex TMTpro (Thermo Fisher) in anhydrous acetonitril (ACN) solution.

\hypertarget{labeling-protocol}{%
\paragraph{Labeling protocol}\label{labeling-protocol}}

\begin{enumerate}
\item Let the TMTpro labels come to room temperature so that all the condensation on the tube is gone. Dissolve the TMTpro reagents in anhydrous acetonitril (ACN) to obtain a concentration of 10 $\mu$g/$\mu$L.
\item Add 1 $\mu$L of TMTpro reagents according to the tables. For one set of plates, you will need 16 $\mu$L of each label. It is recommended to aliquot a slightly larger volume, of around 20 $\mu$L.
\item Seal the plates.
\item Centrifuge for 2 minutes at 120 G.
\item Incubate for 1 hour at room temperature.
\item Prepare a 5% hydroxylamine solution in water.
\item Add 1 $\mu$L of the 5% hydroxylamine solution to each well.
\item Seal the plates.
\item Centrifuge for 2 minutes at 120 G.
\item Incubate for 1 hour at room temperature.
\item Cover the plates with cap strips.
\end{enumerate}

@@

\hypertarget{pooling-protocol}{%
\paragraph{Pooling protocol}\label{pooling-protocol}}

@@

Data was analysed using commercially available hardware, find details below:
Lenovo Laptop X360 with:
Intel(R) Core(TM) i5-6200U CPU @ 2.30GHz
8 GB RAM

@@

{\centering \includegraphics[width=0.8\linewidth]{Thesis_files/figure-latex/data-processing-pipeline-flowchart-vertical-1}

@@

\caption{Flowchart presenting the processing by ProteoScanR}\label{fig:data-processing-pipeline-flowchart-vertical}
 \endfigure\egroup

@@

In order to streamline the analysis of multiple experiments and software development, object oriented programming was applied. The approach in R is to create a so called Q-feature object, which contains all variables and metadata in a hierarchical structure. The structure enables sub setting for further analysis \citep{Vanderaa2021}.

@@

The next step for quality control is the exclusion of spectra with a high false discovery rate (FDR). When applying multiple statistical testing (e.g.~t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant result will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. Many search engines face that problem by calculating the false discovery rate, by dividing false PSMs (=hit of the decoy database) with the total number of PSMs above the peptide-spectrum matching score. The peptides spectrum matching score is defined as \(-10 * log(p)\). Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda \citep{Cox2011}, an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot \citep{Brosch2009}. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

@@

In data science, aggregation refers to a row-wise operation that merges data based on a particular column using a specific function. In the context of processing of PSMs to unique peptides, the desired column is the peptide sequence. To account for different distributions across multiple assays, the median of the channel is used as the function to aggregate multiple matches into one.

@@

Sample size is often a limiting factor in hypothesis testing. A strict quality control and the fact that TMT reagents are only available up to 18-plex can reduce the number of observed samples below the critical threshold, leading to diminished statistical power. To overcome this limitation, the provided software is capable of processing multiple runs simultaneously, allowing for testing of multiple batches and increasing the number of samples that can be included in the analysis. For further processing the runs are joined in one object at the peptide level.

@@

Columns which do not meet the desired intensity can be filtered by a threshold set on the RI. The median RI can also be used to check if an entire channel has a lower detection level. This can be due two reasons. One is the expression level of the given proteins in a sample. Meaning, that the expression of the observed sample type is simply lower than in the other type. Another one could be a spillage of TMT detection in other channels due incorrect or missing correction of the TMT isotopes. The correction values can be found on the vendor`s webpage (in this case Thermo Fisher) and correction needs to be set up in MaxQuant.

@@

Depending having a bulk sample or single-cell sample choosing a minimum of observed peptides and a cutoff value for the CV, changes the level of confidence in the peptide data. The coefficient of variation of a peptide is defined as the ratio of the standard deviation to the mean and describes the relationship of the observed peptide signal over multiple proteins (=razor proteins). Peptides having a high coefficient of variation over many razor proteins are considered as noise and need to be filtered out before statistical analysis.

@@

Data transformation applies a function to each value of a matrix or array, so that:
\[
y_i = f(x_i)
\]
Depending on the distribution of the values in the observed expression set, different transformations can be applied to fulfill assumptions for statistical testing. In the shiny application, various procedures are implemented and can be further expanded upon request from the user. The macrophage analysis done by Specht et al.~mentioned in their SCoPE2 publication \citep{Specht2021} uses the logarithm to the base 2 to spread a compacted distribution and remove skewness in the dataset. This transformation was used as a reference when comparing methods. However, using the logarithm to the base 10 has the same effect and improves the interpretability of the values. Furthermore the log10 transformation will also be facilitated by other transformation methods such as the boxcox method, which is also implemented in the application.
When observing a wide distribution of small and large values of the expression set in a histogram, a square root transformation can help increase the variability of smaller values and decrease the variability of larger values.
Depending on the distribution, the user has to decide which transformation to consider and verify that statistical assumptions are met after application with visualizations such as histograms, qq-plots, and MA-plots.
For statistically inexperienced users, the boxcox transformation can help with this decision. Developed in 1964 by Box and Cox \citep{Sakia1992}, this method applies a linear model against 1 on the data to determine the statistical parameter lambda by the maximum likelihood method. The log-likelihood method takes the 95\% confidence interval for the parameter lambda and the final lambda is chosen as the value with the highest log-likelihood value.

@@

The term normalization is ambiguous in data-science and will be explained briefly in this chapter. Starting with converting the data to the Z-distribution (Gaussian result), where 0 represents the mean and 1 represents the standard deviation, in vectorization every feature (or protein) is represented as a vector that points in a specific direction in the unit sphere, normalization can have different meanings. Normalization is considered as part of the scaling procedure and ensures that every feature contributes equally to the statistical model and hinders large values from biasing the model in a particular direction. However, this could also have negative consequences for the modeling procedure, as it may reduce the impact of important features on the dataset.
Depending on the algorithms later used in the data processing, the method of choice for normalization can change the results significantly. The K-nearest neighbor (K-NN) clustering algorithm, for instance relies on the distances between distinct data points by minimizing variance of the squared euclidean distances. This method will be applied in the missing value imputation, which will be the next step applied to the data.
The first normalization method available in the user interface is column-wise median and row-wise mean normalization. This method was chosen by Specht et al in the SCoPE2 publication 2021 \citep{Specht2021} and worked as a reference during the development process to benchmark other methods. When dividing each value of an expression set by the median of the particular column, the procedure is considered as a column-wise median normalization. Although the mean-normalization works the same way but takes the mean as the second variable. In the analysis pipeline both procedures were applied on the dataset. However when the logarithm as the transformation method was chosen as transformation method, ProteoScanR automatically switches from a division to a subtraction, since:
\(\log_a \left(\frac{u}{v}\right)=\log_a u-\log_a v\). After applying this normalization method, each value can be interpreted as a distance or fraction of the mean or median of the corresponding column or row, enabling fair comparisons between features and samples.
The second method for normalization included is the CONSTANd method, which was proven suitable for relative quantification in the field of proteomics \citep{Maes2016, VanHoutven2021}, but can also be applied on RNAseq data. CONSTANd uses a technique called matrix raking, which employs the RAS (\(K=RAS\)) algorithm. The expression set reflects the non-negative real \((m, n)\) matrix \(A\) were the bi proportional constrained matrix problem will be solved by finding the \((m, n)\) matrix \(K\) which equals to \(diag(x) * A *diag(y)\), so that \(K=diag(x) * A *diag(y)= RAS\). Whereas \(x \in \mathbb{R}^{m}\), \(y \in \mathbb{R}^{n}\) and \(R\) denotes for the row multiplier and \(S\) for the column multiplier. The solution to this problem involves finding the row sum of \(K=u_i\) and the column sum of \(K=v_j\), where \(i\) and \(j\) denote the row and column indices of the matrix, respectively \citep{Bacharach1965}. In other words, the matrix will be alternatingly manipulated on rows and columns until the mean of both equals 1. In order to apply the method and yield true results, one has to consider 3 major assumptions between sample types. These assumption can be observed in the MA-plot which indicates the differences of two samples. Given the intensities of two samples, \(R\) and \(G\), visualized on a graph with the x-axis as \(M=\log_2(\frac{R}{G})\) and the y-axis as \(A=\frac{1}{2}\log_2(RG)\). A comparison between all sample types of identically processed sets is needed before calculation and the assumptions must be fulfilled.

@@

When analyzing multiple batches simultaneously, it is possible to encounter situations where a particular protein is not detected in one of the batches. In order to handle this dark space in the expression set, several methods can be chosen in the application.
A variety of function employs replacing the missing values with the mean or median of the rest of the matrix (total mean/median). This procedure introduces no bias in the data if the missingness is low, although differential expression for the sample carrying the missing feature can not be expected as well. An even more conservative approach is dropping the rows for the feature with the missing value in one of the observed samples.
A common practice in proteomics is utilizing the K-nearest neighbor algorithm (=KNN), which imputes the missing values according to an euclidean metric of the neighboring values in the columns where the feature is not missing. \(K\) denotes for the count of neighboring values of a protein which are considered and can be selected upon preference. The euclidean distance between two points is defined as \(d = \sqrt{a^2 * b^2}\), the \(K\) nearest neighbors average value will be assigned to the intensity of the missing feature. Since the average is a statistical metric which can be biased easily by skewed distributions, normalization is an important task before performing missing value imputation. KNN was benchmarked for the regression of protein abundance intensities and showed an Area under the Curve (AUC) above 0.7 when testing on a human proteome test set \citep{Lan2013}.

@@

After differential expression the upcoming question could be, in which context the up or down regulated proteins are. Enriched pathways can be understood as an over-representation (ORA = over-representation analysis) of proteins within a specific set. This set, also known as ontology can be a pathway or a selection of common denominators in biological contexts, such as cell types or oncological factors. The statistical testing to identify the overabundant proteins of a specific set is similar to a \(\chi^2\) test developed by Karl Pearson \citep{Pearson1900} or the exact Fisher test \citep{Sprent2011} for tables with 4 fields in small sample sizes. Within the list are all truly differential expressed genes, an annotation is the identification of a protein in the desired biological context.
The table is expected to be populated by the values as follows:

@@

Under the null hypothesis \(H_{0}\) states that the distribution and population of the table is random and there is no clear tendency with significant evidence. Against that the alternative hypothesis \(H_{A}\) says that there is an underlying tendency the table is populated.
This leads to the test statistics, the observed values (=\(O\)) in the table are significantly different to the expected (=\(E=\frac{(A+B \lor C+D)}{N}\) ones. One approach is approximating \(\chi^2= \sum_{i=1}^{n} \frac{(O_i - E_i)^2}{E_i}\) and deriving the p-value by the area under the density curve of the \(\chi^2\) distribution \(Q = \sum_{i=1}^{k} Z_{i}^{2}\). In the distribution \(k\) is defined by the degrees of freedom (=df), what can be obtained by \(k=(columns-1)*(rows-1)=df\) of the table. Another approach to obtain an exact p-value would be Fisher`s exact test, which states that the margins are distributed according to the hypergeometric distributions. The test statistic leads directly to the p-value with no need for approximation:
\(p=1-\sum_{i=0}^{A-1}\frac{\binom{B}{i}\binom{D-B}{A+C-i}}{\binom{D}{A+C}}\)
Here D denotes either for the background distribution which can be the total number of proteins obtained in the experiment, all proteins with annotation or a custom selected background. The background indicates all proteins which could be positive tested for in the observed sample. The above mentioned three options are implemented in the user interface.
After obtaining the p-values for each protein set, the subsequent critical step involves the correction for multiple testing, considering the inherent risk of false positive results. Multiple testing correction methods adjust the significance thresholds to account for the increased probability of detecting false positives when performing multiple statistical tests. In this protein set enrichment analysis, the user is provided with a selection of established correction methods, akin to those commonly utilized in the field of differential expression analysis, ensuring robust and reliable interpretation of the results.
Once the enriched protein pathways are determined, they can be visually represented using bar or dotplots, and a network of the found pathways.

@@

For custom ontologies the protein enrichment analysis is done using the R bioconductor package piano \citep{Vaeremo2013}. In opposite to the pathway enrichment, piano maps proteins to a pre-selected gene collection set, which can be either downloaded from gsea-msigdb.org or custom made for the particular experiment and explains conditions, phenotype or other desired properties. The online database offers pre-built gene set collections for previous research questions in multiple fields of human biology.
For statistical testing the above explained Fisher exact test is used. The output of the Fisher test does not show any directionality in regards of protein expression. If the user wants to include directionality in the network diagram, the default option of piano can be included in future versions of the program.

@@

To begin using the application, the first step is to upload the tab-delimited evidence file and the sample annotation file after processing the data with MaxQuant. Once these files are uploaded, the application is ready to run with default values. The default settings are preconfigured for an initial analysis, and any further adjustments or fine-tuning can be made in subsequent steps of the analysis process.

\hypertarget{pre-processing-and-quality-control}{%
\subsection{Pre-processing and quality control}\label{pre-processing-and-quality-control}}

The initial stages of the pipeline are focused on creating a reliable dataset for analysis that is free from noise and bias.

@@

\caption{Summary showing the processing steps of the analysis pipeline}\label{fig:ui-summary}
 \endfigure\egroup
